Are historical stage records useful to decrease the uncertainty of flood frequency analysis ? A 200-year long case study

### 2022-12-08 ArticleChProp/main.tex
Corrections Jerome 90%

\title{Are historical stage records useful to decrease the uncertainty of FFA? A 200-year long case study}
\author{Mathieu LUCAS, Benjamin RENARD, Jérôme LE COZ, Michel LANG, Gilles PIERREFEU}

@@

    This work proposes methods to make the most of long discharge series in flood frequency analysis, and provides insights on how historical data may or may not improve design floods estimates. Flood frequency analysis based on peak flow series from hydrometric stations is widely used to estimate flood hazard. The uncertainty around flood frequency estimates depends on sample size and on the uncertainty of streamflow data. Documentary evidence allow lengthening the continuous discharge series available and therefore has the potential to reduce sampling uncertainty. However, specific uncertainties affect those data as the stage/discharge relationship is generally difficult to estimate in this century-long context and as ancient visual stage reading were less frequent than recent sensor-measured stages. A consistent treatment of uncertainty over the whole data processing chain seems essential but is usually not performed. In the proposed approach, uncertainties are estimated and propagated from stage measurements to extreme flood quantiles estimates using Monte Carlo procedures. This allows exploring the role of both streamflow and sampling uncertainties in design floods estimations. This approached is applied to an exceptional 205-year long continuous discharge series of the Rhône River at Beaucaire. Results show that sampling uncertainty is reduced when sample size increases from 50 to 100 years, but is not much reduced beyond 100 years. However, the streamflow uncertainty contribution is increased when considering older and thus more uncertain floods.

@@

\section{Introduction}
    \paragraph{}
    Flood frequency analysis (FFA) is a widely used method to estimate flood hazard. It allows linking the magnitude of a flood to its probability of occurrence (\citet{hamed_flood_2019}; \citet{jain_design_2019}). Design floods, estimated for various exceedance probabilities (or equivalently, return periods), are commonly used for population safety policies, land use planning, as well as industrial safety. The most standard FFA approach is to estimate a distribution using a sample of flood peaks (typically, annual maximum discharges or discharges over a given threshold). Extrapolation of this distribution allows reaching the desired flood quantile that commonly corresponds to a 100 or 1000-year return period. 
    % refs FFA from FFA book ?

    \paragraph{}
    This FFA approach is affected by various sources of uncertainty. First, the hydrological data used to estimate the FFA distribution is uncertain: indeed, the usual process for deriving streamflow time series is going through steps that lead to measurement (gaugings and stage) and model (rating curve) errors. Moreover, the estimated FFA distribution is itself uncertain because of the limited size of the available streamflow data \citep{kjeldsen_uncertainty_2011}. Considering the decisions that rely on flood frequency analysis, a consistent treatment of uncertainty all over the data processing chain is essential but is usually not performed.
    % more references here again

    \paragraph{}
    Streamflow series are affected by several sources of uncertainty as described by \citet{mcmillan_benchmarking_2012} benchmark. First, stage measurements are affected by various sources of uncertainty. This uncertainty is often neglected but may have a important impact on the final hydrograph \citet{hamilton_quantifying_2012}, especially for historical stage series. A large number of stage error sources is identified in the literature (\citet{van_der_made_determination_1982}; \citet{petersen-overleir_uncertainty_2005}; \citet{mcmillan_benchmarking_2012}, \citet{horner_impact_2018}), such as staff gauge reading errors, levelling of the staff gauge, or stage sensor calibration. The frequency of measurement may also induce errors. With modern automatic systems, stage is measured with a time step generally smaller than the speed of variation of the stage (e.g. between 15 min and 1 hour), which leads to negligible errors. However, before the rise of automatic systems, measurements were made by operators who were reading the staff gauge less frequently (e.g. once or a few times per day), thus possibly missing the flood peak. This issue is particularly important when old series are used. To estimate the measurement frequency uncertainty, \citet{hamilton_quantifying_2012} and \citet{kuentz_hydrometrie_2014} analysed this situation by sub-sampling recent data for which sub-hourly measurements are available. This allows calculating the difference between the variable of interest (such as the daily maximum stage) derived from scarce data, and the same variable derived from high-frequency measurements. \citet{kuentz_hydrometrie_2014} applied the calculated bias to correct old stage series at a monthly time step. This correction aimed at taking into account the error due to the daily variability caused by snow melt during cold months. However, this type of correction have never been applied to peak stage correction during floods, especially in the case of long stage series.

@@

    Rating curve uncertainty is also a main issue when dealing with streamflow series. Transforming stage into discharge requires calibration data (gaugings) to establish the stage/discharge relationship. Gaugings uncertainty depends on the method used (\citet{lecoz_quantification_2014} and \citet{puechberty_charte_2017}). Moreover, the rating curve is also affected by uncertainties coming from the imperfection of the chosen model to represent the actual hydraulic configuration and from parameter estimation. Many methods have been proposed to quantify these uncertainties (\citet{petersen-overleir_bayesian_2009}; \citet{juston_rating_2014}; \citet{le_coz_combining_2014}; \citet{morlot_dynamic_2014}; \citet{coxon_novel_2015}; \citet{mcmillan_rating_2015}; \citet{mansanarez_rapid_2019}). A comparison of several of these methods has been recently proposed by \citet{kiang_comparison_2018}. Another important issue affecting streamflow data accuracy is rating changes. The stage/discharge relationship is frequently affected by changes coming from various origins that can be natural or anthropic, for instance: bed geometry evolution during floods or river works, aquatic vegetation growth and decay, ice sheets... A regular monitoring through gaugings is essential to detect those changes \citep{ibbitt_gauging_1987}, that can be transient or sudden. Several methods have been proposed to deal with rating changes: estimating rating curves on moving temporal windows (\citet{westerberg_stage-discharge_2011} and \citet{guerrero_temporal_2012}), computing as many rating curves as there are gaugings \citep{morlot_dynamic_2014}, exploring changes in the annual minimum stages \citep{lapuszek_methods_2015}, selecting 0.5-years return period discharge as a threshold for rating changes \citep{mcmillan_impacts_2010}... More recently, \citet{darienzo_detection_2021} proposed a method based on a recursive segmentation procedure, accounting for both gaugings and rating curve uncertainties. This method has a particular interest when dealing with ancient and thus uncertain gaugings. Following the detection of rating shifts, rating curves should be estimated for each stable sub-period. This task may not be straightforward, as the number of gaugings available within a sub-period is not always sufficient to properly estimate the stage/discharge relationship for the whole discharge range. A common way to address this problem is to artificially repeat some gaugings from other stability periods (\citet{mcmillan_benchmarking_2012}; \citet{puechberty_charte_2017}). \citet{mansanarez_shift_2019} proposed an alternative approach to deal with this issue. They developed a stage-period-discharge (SPD) model  where some parameters of the rating curve are allowed to vary across periods, while the others are supposed fixed. This method has the advantage of transferring information between periods to improve the rating curve estimation for all the stable stage/discharge periods, even when few gaugings are available.

@@

    Estimating sampling uncertainty in FFA is a well-established approach. Whatever the chosen distribution and estimation method, standard statistical procedures are available. However, these standard procedures only quantify sampling uncertainty, and do not consider uncertainties affecting the data. The literature review proposed in the previous paragraphs show that methods for quantifying individual sources of uncertainty (stage, rating curve and FFA distribution estimation) are available. However, the way all these uncertainties propagate through the FFA analysis chain has been less thoroughly studied. A few solutions have emerged to propagate uncertainties in stage time series through uncertain rating curves (Dymond1982, Herschy 1998, \citet{petersen-overleir_uncertainty_2005}), but they assume independent stage errors and therefore neglects systematic errors. \citet{horner_impact_2018} proposed a consistent method for the propagation of both sources of stage uncertainty through uncertain rating curves. Therefore, it is possible to infer the contribution of each source of uncertainty in streamflow data. \citet{petersen-overleir_accounting_2009}, \citet{steinbakk_propagation_2016}, and \citet{vieira_assessing_2022} performed an integrated analysis where the rating curve parameters and the flood frequency distribution are both estimated. All these studies highlighted the importance of considering rating curve uncertainty for design flood estimations and concluded that, under some conditions, accounting for rating curve uncertainty may notably widen the uncertainty intervals around flood quantiles. However, none of them considered both stage measurement errors and rating changes, which may constitute a major source of uncertainty for streamflow data. This is particularly the case when dealing with long streamflow series for which stage uncertainty is important and rating changes may have been missed. Their consideration in a flood frequency framework is therefore not avoidable. 

@@

    \paragraph{}
    The following questions will be considered in this paper: 
    \begin{itemize}
        \item[1.] How to make the most of historical hydrometric data in flood frequency analysis while accounting for uncertainties at each step of the procedure ?

@@

    \paragraph{}
    This paper will illustrate the chained application of methods to quantify and propagate uncertainty from stage records (and their limited time resolution) and stage-discharge rating curves to the estimation of extreme flood quantiles. While most of these methods already exist, a key novelty of this work is their combination to provide an end-to-end evaluation of the  uncertainty affecting FFA estimations, from stage to flood quantiles. An original method to quantify the stage uncertainty stemming from infrequent readings is also proposed.

    The paper is organized as follows. First, the methodology for estimating uncertain streamflow series in a century-long context is presented (section 2.1). It goes through the detection of rating shifts (section 2.2), the estimation of rating curves (section 2.3), and the estimation and propagation of stage errors (section 2.4). Then, an approach to propagate streamflow uncertainty through the estimation of extreme flood quantiles is proposed (section 2.5). This procedure is illustrated through its application to the Beaucaire gauge on the Rhône River (section 3), for which official flood hazard assessments only used discharge series from 1920 \citep{rigaudiere_etude_2000}. The recent works of \citet{pichard_hydro-climatology_2017} and \citet{bard_actualisation_2018} allowed building a continuous discharge series from 1816 to the present time, which makes it the ideal case study for this procedure. The results of this application are presented in section 4, and they are discussed in section 5, where avenues for improvements are proposed.

@@

    The uncertainty propagation framework (figure \ref{fig:ChProp}) is a chain of uncertainty estimation methods. It starts with the basic hydrometric data: stage series and gaugings, and the final results are the flood quantiles with uncertainty propagated from each previous step of the procedure. This uncertainty propagation is achieved by Monte Carlo methods.

    The first three steps of the procedure rely on a rating curve estimation method. The BaRatin method \citep{le_coz_combining_2014} is used in this work. The Bayesian paradigm is used to estimate the parameters of the general rating curve equation, a combination of power equations: $Q = a(h-b)^c$, where Q is the discharge in m\textsuperscript{3}/s, the flow depth is the difference between the stage h (in m) and an offset b, and a and c being the coefficient and the exponent of the power function. The rating curve equation is deduced from a hydraulic analysis of the gauging station, aimed at identifying the main hydraulic controls governing the stage-discharge relation. The different controls can be activated successively or simultaneously. As an example, in a two-control hydraulic configuration, the rating curve $Q(h)$ may be expressed as:

    \begin{itemize}
        \item $k_1 < h \leq k_2 $ (main channel): $Q(h) = a_1(h-b_1)^{c1}$
        \item $h \geq k_2$ (main channel + floodway):  $Q(h) = a_1(h-b_1)^{c1}+ a_2(h-b_2)^{c2}$
    \end{itemize}

    Parameters $a$, $b$ and $c$ can be linked with the physical properties of the control. For instance for a channel control with a wide rectangular cross-section, using a Manning-Strickler equation, we have: $a=KB\sqrt{S}$  ; $b=z$ ; $c=5/3$

    Bayesian inference allows deriving the posterior distribution of rating curve parameters by combining hydraulic information (priors for parameters of each hydraulic controls) and information from gaugings with uncertainty (likelihood). The posterior distribution is explored with a Markov Chain Monte Carlo (MCMC) sampler \citep{renard_application_2006}. Two sources of uncertainty are associated with the estimated rating curves. Parametric uncertainty reflects the uncertainty due to the rating curve parameters estimation because of the limited amount of gaugings and the gaugings uncertainty. Remnant uncertainty comes from the imperfection of the chosen rating curve model to represent the actual hydraulic configuration. We refer the reader to \citet{le_coz_combining_2014} for a more thorough description.

@@

    The stage/discharge relationship is sensitive to sudden changes caused by morphogenic floods or other causes affecting the flow characteristics. Relying on residuals between gaugings and rating curve is the most common approach to monitor the stability of this relationship over time. The method proposed by \citet{darienzo_detection_2021} is used in this work. First, a baseline rating curve is estimated using the whole gaugings dataset. The residuals between gaugings and the rating curve are determined, and a statistical segmentation test is applied to those values. This segmentation procedure accounts for the residuals uncertainty, coming from both the gaugings and the rating curve uncertainties. The optimal number of segments is determined according the Bayesian Information criterion (BIC). Then, the same steps are applied recursively to each sub-period determined earlier. The recursive procedure is stopped when the BIC indicates that a single period is optimal. The advantage of this method is that both gaugings and rating-curve uncertainties are accounted for in a rigorous Bayesian framework. The results are not only the dates of rating-shifts but the posterior probability density function (pdf) of change point time. This allows affecting the shift time to the time of the maximum stage included in the posterior credibility interval. Prior knowledge is provided on the expected magnitude residuals, the number of segments at each iteration and the minimum time lag between two consecutive change points.
    %paramétrage et les options que tu retiens pour la suite de l'article

@@

    \paragraph{}
    Once the stability periods have been identified, the next step is to estimate the rating curve associated with each period. \citet{mansanarez_shift_2019} developed a stage-period-discharge (SPD) model "based on the physical interpretation of changes in the stage-discharge relation across a series of stability periods". Prior knowledge is provided on the hydraulic controls and on the rating changes (i.e. estimated amplitude of change between the rating curves of two successive periods). In this model, some parameters of the rating curve equation are supposed to vary across the periods while the others do not. The BaRatin SPD method allows transferring information not only between periods, but also between controls, taking into account the uncertainty of both gaugings and of the rating curve model. This can be achieved without the need to artificially repeat gaugings across periods, which is of particular interest for stations with scarce gaugings. The identification of the potentially varying parameters is based on an hydraulic analysis of the site and and is an important step. Generally, changes are suspected coming from offsets or channel width parameters. The so-called local changes are supposed to affect the lowest control only (for instance the movement of the controlling riffle), whereas the global changes are affecting various controls at the same time (for instance, the scouring or filling of the channel, affecting low flows and main channel offsets at the same time). See \citet{mansanarez_shift_2019} for a detailed description of prior specification for time-varying rating curves.

@@

    Many sources of error having distinct statistical properties can affect stage measurements, as described in \citet{horner_impact_2018}. Five different sources of uncertainty ($\delta_{1,...,5}$) affecting stage measurements are considered. Let $h(t)$ be the measured maximum stage of a day $t$. $\hbar(t)$, the unknown true maximum stage is assumed to be approximated by the following equation:

@@

    For which: 
    \begin{itemize}

        \item[1 -]Staff gauge reading errors $\delta_1 \sim \mathcal{N}(0,\mu_1)$ originate from operators reading the gauge, where $\mu_1$ is depending on the precision of graduations (usually 1cm), and can be aggravated by the presence of waves, especially during floods \citep{mcmillan_benchmarking_2012}.

        \item[-]Most stage measurements nowadays are with automatic sensors of various types such as pressure sensors, floaters, radars, and they require a calibration to link the water stage to the measured proxy (respectively the pressure of the water column, the height of a floater, or the air draught). Two types of errors arise from this process: 
        \begin{itemize}
            \item[2 -] Sensor errors $\delta_2 \sim \mathcal{N}(0,\mu_2)$, where $\mu_2$ being usually estimated by the sensor constructor.

            \item[3 -] Sensor calibration errors $\delta_3 \sim \mathcal{N}(0,\mu_3)$ are related to the corrections made by operators when comparing the proxy measured by the sensor to the current stage at the staff gauge reference. An operator error at this step could affect the stage measurement until the next calibration. Sensor calibration error $\delta_3$ is hence constant between two calibrations and can be represented by drawing a new random value at each operator intervention.
        \end{itemize}

        \item[4 -]Datum errors $\delta_4  \sim \mathcal{N}(0,\mu_4)$ are related to changes in the datum reference measurement (usually the staff gauge zero value) and possible mismatches between successive gauges. Similarly to $\delta_3$, this error is constant between two gauge changes or datum reference measurements.

        \item[5 -] Measurement frequency errors $\delta_5$ are related to the inadequacy of the frequency of measurement with respect to the speed of stage variations. As explained in the introduction, ancient stage measurements were done by operators who were reading the staff gauge much less frequently than modern automatic systems, which may lead to missing and hence underestimating the peak stage during floods. Unlike other types of stage errors, this error is hence necessarily positive, which calls for using a positive distribution such as the Log-normal or Exponential distribution at each measurement time-step. The parameters of this distribution can be estimated with data from the recent period, by analyzing the difference between the daily maximum stage derived from the high-frequency sensor measurement and that from an infrequent fixed-time reading. Note that the measurement frequency error for hourly (or less) measurements are considered negligible when considering large rivers with slow variations.

    \end{itemize}

    \paragraph{}
    To sum up, $\delta_1$, $\delta_2$ and $\delta_5$ errors are drawn at each time step, while $\delta_3$ and $\delta_4$ errors are only drew at specific calibration times. $\delta_1$ to $\delta_4$ are assumed Gaussian with known standard deviations, while $\delta_5$ is assumed Exponential with parameter estimated by subsampling recent measurements. Monte Carlo methods are applied to combine these multiple uncertainty sources. The uncertainty around stage measurements from eq. \ref{eq:StageError} is hence represented by 500 possible realisations of the stage $h(t)$.

@@

   Stage measurements realisations can be propagated through uncertain rating curves, following the approach described by \citet{horner_impact_2018}. To estimate the contribution of the different sources of streamflow uncertainties, four scenarii are considered:

   \begin{itemize}
       \item \textbf{Case 1: Maxpost streamflow.} Stage measurements are free of uncertainties and are assumed to be the median of the stage time series realisations. Rating curve estimation is also free of uncertainties. The unique stage time series is propagated trough the maxpost rating curves, resulting in one set of discharges time series. 
       \item \textbf{Case 2: Stage uncertainty.} Stage measurements are uncertain. This uncertainty is represented by $n$ possible stage time series, propagated through the maxpost rating curves. $n$ sets of discharge time series are obtained.
       \item \textbf{Case 3: Stage and parametric rating curve uncertainty.} Stage measurements are uncertain and rating curve estimation is affected by parametric uncertainties (represented by $m$ sets of rating curves) but is not affected by remnant uncertainties. The $n$ stage time series are propagated through $m$ sets of rating curves, leading to $n$ x $m$ sets of discharge time series. 
       \item \textbf{Case 4: Total streamflow uncertainty.} Stage measurements are uncertain and rating curve estimation is affected by both parametric and remnant uncertainties, represented by $m$ sets of rating curves. The $n$ stage time series are propagated through $m$ sets of rating curves, leading to $n$ x $m$ sets of discharge time series that represent the total streamflow uncertainty.
   \end{itemize}

@@

    \subsubsection{Pont de Beaucaire}

>>>>

\title{Are historical stage records useful to decrease the uncertainty of flood frequency analysis ? A 200-year long case study}
\author{Mathieu LUCAS, Benjamin RENARD, Jérôme LE COZ, Michel LANG, A. BARD, G. PIERREFEU}

@@

    This paper presents a chain of uncertainty estimation methods for flood frequency analysis in the context of long discharge series. Uncertainties are estimated and propagated from stage and rating curves to extreme flood quantiles estimates using Monte Carlo procedures. This method allows exploring the role of both streamflow and sampling uncertainties in design floods estimations and provides insights on how historical data may or may not improve design floods estimates. The procedure is applied to the 205-year long continuous discharge series of the Rhône River at Beaucaire (95 590 m\textsuperscript{2}). Total uncertainty on flood quantile is significantly reduced when the length of the series increases from 50 to 100 years. Beyond this length, the total uncertainty remains stable, despite 19th century flood discharges have larger uncertainties than “modern’ flood discharges (20th century and present). Addition of the two largest known floods in 1840 and 1856 remains critical, as the estimate of Q100 and Q1000 floods is revised upwards.

@@

\section{Introduction}
    \paragraph{}
    Flood frequency analysis (FFA) is a widely used method to estimate flood hazard. It allows linking the magnitude of a flood to its probability of occurrence (\citet{hamed_flood_2019}; \citet{jain_design_2019}). Design floods, estimated for various exceedance probabilities, or equivalently, return periods, are commonly used for population safety policies, land use planning, as well as industrial safety. The most standard FFA approach is to estimate a distribution using a sample of flood peaks, typically, annual maximum discharges or discharges over a given threshold. This distribution may be extrapolated to reach the desired flood quantile that commonly corresponds to a 100 or 1000-year return period. 
    % refs FFA from FFA book ?

    \paragraph{}
    This FFA approach is affected by various sources of uncertainty. First, the hydrological data used to estimate the FFA distribution is uncertain. Indeed,
    streamflow time series are generally derived from stage time series through rating curve models. This procedure leads to measurement (gaugings and stage) and model (rating curve) errors. Moreover, the estimated FFA distribution is also uncertain because of the limited size of the available streamflow data \citep{kjeldsen_uncertainty_2011}. Considering the decisions that rely on FFA results, a consistent treatment of uncertainty all over the data processing chain is essential but is usually not performed.
    % more references here again

    \paragraph{}
    Streamflow series are affected by several sources of uncertainty as described by \citet{mcmillan_benchmarking_2012} benchmark. First, stage measurements are affected by various sources of uncertainty \citep{horner_impact_2018}. This uncertainty is often neglected but may have a substantial impact on the final discharge series \citet{hamilton_quantifying_2012}, especially for historical stage series. A large number of stage error sources is identified in the literature (\citet{van_der_made_determination_1982}; \citet{petersen-overleir_uncertainty_2005}; \citet{mcmillan_benchmarking_2012}, \citet{horner_impact_2018}), such as staff gauge reading errors, levelling of the staff gauge, or stage sensor calibration. The frequency of measurement may also induce time interpolation errors. With modern automatic gauges, stage is generally measured with a small enough time step by respect to the speed of variation of the stage (e.g. between 15 min and 1 hour), to get negligible time interpolation errors. However, before the rise of automatic gauges, measurements were made by operators who read the staff gauge less frequently (e.g. once or a few times per day), thus possibly missing the flood peak. This issue is particularly important when old stage series are used. \citet{hamilton_quantifying_2012} and \citet{kuentz_hydrometrie_2014} estimated the measurement frequency error by sub-sampling recent, sub-hourly measurements. They calculated the difference between the variable of interest (such as the daily maximum stage) derived from scarce data, and the same variable derived from high-frequency measurements. \citet{kuentz_hydrometrie_2014} applied the (monthly averaged) calculated bias to correct old stage series. This correction aimed at taking into account the error due to the daily variability caused by snow melt during cold months. However, this type of correction has never been applied to peak stage correction during floods, especially in the case of long stage series.

@@

    Rating curve uncertainty is also a major issue when dealing with streamflow series. Transforming stage into discharge requires calibration data (gaugings) to establish the stage/discharge relationship. Gaugings uncertainty depends on the measurement method (\citet{lecoz_quantification_2014}b and \citet{puechberty_charte_2017}). Moreover, the rating curve is also affected by uncertainties coming from the imperfection of the chosen model to represent the actual hydraulic configuration and from parameter estimation. Many methods have been proposed to quantify these uncertainties (\citet{petersen-overleir_bayesian_2009}; \citet{juston_rating_2014}; \citet{le_coz_combining_2014}; \citet{morlot_dynamic_2014}; \citet{coxon_novel_2015}; \citet{mcmillan_rating_2015}; \citet{mansanarez_rapid_2019}). A comparison of several of these methods has been recently proposed by \citet{kiang_comparison_2018}. Another important issue affecting streamflow data accuracy is rating changes. The stage/discharge relationship is frequently affected by changes coming from various origins that can be natural or anthropic, for instance: bed geometry evolution during floods or river works, aquatic vegetation growth and decay, ice cover... A regular monitoring through gaugings is essential to detect those changes \citep{ibbitt_gauging_1987}, that can be transient or sudden. Several methods have been proposed to deal with rating changes: estimating rating curves on moving temporal windows (\citet{westerberg_stage-discharge_2011} and \citet{guerrero_temporal_2012}), computing as many rating curves as there are gaugings \citep{morlot_dynamic_2014}, exploring changes in the annual minimum stages \citep{lapuszek_methods_2015}, selecting 0.5-years return period discharge as a threshold for rating changes \citep{mcmillan_impacts_2010}. More recently, \citet{darienzo_detection_2021} proposed a method based on a recursive segmentation procedure, accounting for both gaugings and rating curve uncertainties. This method has a particular interest when dealing with old and thus uncertain gaugings. Following the detection of rating shifts, rating curves should be estimated for each stability period. This task may not be straightforward, as the number of gaugings available within a stability period is not always sufficient to properly estimate the stage/discharge relationship for the whole discharge range. A common way to address this problem is to artificially repeat some gaugings from other stability periods (\citet{mcmillan_benchmarking_2012}; \citet{puechberty_charte_2017}). \citet{mansanarez_shift_2019} proposed an alternative approach to deal with this issue. They developed a stage-period-discharge (SPD) model  where some parameters of the rating curve are allowed to vary across periods, while the others are supposed fixed. This method has the advantage of transferring information between periods to improve the rating curve estimation for all the stable stage/discharge periods, even when few gaugings are available.

@@

Estimating sampling uncertainty in FFA is a well-established approach. Whatever     the chosen distribution and estimation method, standard statistical procedures are available \citep{beven_applied_2014}. However, these standard procedures only quantify sampling uncertainty, they do not consider data uncertainty. The literature review proposed in the previous paragraphs shows that methods for quantifying individual sources of uncertainty (stage, rating curve and FFA distribution estimation) are available. However, the way these multiple uncertainties propagate through the FFA analysis chain has been less thoroughly studied. A few solutions have emerged to propagate uncertainties in stage time series through uncertain rating curves (\citet{dymond_accuracy_1982}, \citet{herschy_hydrometry_1998}, \citet{petersen-overleir_uncertainty_2005}), but they assume independent stage errors and therefore neglect systematic errors. \citet{horner_impact_2018} proposed a consistent method for the propagation of both sources of stage uncertainty through uncertain rating curves. Therefore, it is possible to infer the contribution of each source of uncertainty in streamflow data. \citet{petersen-overleir_accounting_2009}, \citet{steinbakk_propagation_2016}, and \citet{vieira_assessing_2022} performed an integrated analysis in which both rating curve parameters and flood frequency distribution are estimated. All these studies highlighted the importance of considering rating curve uncertainty for design flood estimations and concluded that, under some conditions, accounting for rating curve uncertainty may notably widen the uncertainty intervals around flood quantiles. However, none of them considered stage measurement and time interpolation errors, as well as rating changes, which may constitute a major source of uncertainty for streamflow data. This is particularly the case when dealing with long streamflow series for which stage uncertainty is large and variable through time, and rating changes may have been missed. Their consideration in a flood frequency framework is therefore not avoidable. 

@@

    \paragraph{}
    The following questions will be considered in this paper: 
    \begin{itemize}
        \item[1.] How to make the most of historical hydrometric data in flood frequency analysis while accounting for multiple and variable uncertainties at each step of the procedure ?

@@

    \paragraph{}
    This paper illustrates the chained application of methods to quantify and propagate uncertainty from stage records (and their limited time resolution) and stage-discharge rating curves to the estimation of extreme flood quantiles (Figure \ref{fig:ChProp}). While most of these methods already exist, a key novelty of this work is their combination to provide an end-to-end evaluation of the  uncertainty affecting FFA estimations, from stage to flood quantiles. An original method to quantify the stage uncertainty stemming from infrequent readings is also proposed.

    The paper is organized as follows. First, the methodology for establishing uncertain streamflow series in a century-long context is presented (section 2.1). It goes through the detection of rating shifts (section 2.2), the estimation of rating curves (section 2.3), and the estimation and propagation of stage errors (section 2.4). Then, an approach to propagate streamflow uncertainty through the estimation of extreme flood quantiles is proposed (section 2.5). This procedure is illustrated through its application to the Beaucaire gauge on the Rhône River (section 3), for which official flood hazard assessments only used discharge series from 1920 \citep{rigaudiere_etude_2000}. The recent works of \citet{pichard_hydro-climatology_2017} and \citet{bard_actualisation_2018} provided a continuous stage series from 1816 to the present time, which makes it the ideal case study for demonstrating this procedure. The results of this application are presented in section 4, and they are discussed in section 5, where avenues for improvements are proposed.

@@

    % The uncertainty propagation framework (figure \ref{fig:ChProp}) is a chain of uncertainty estimation methods. It starts with the basic hydrometric data: stage series and gaugings, and the final results are the flood quantiles with uncertainty propagated from each previous step of the procedure. This uncertainty propagation is achieved by Monte Carlo methods.

    % \begin{figure}[h!]
    %     \centering
    %     \includegraphics[width = 8cm]{Figs/1-uTotSchema.png}
    %     \caption{Block diagram of the uncertainty propagation procedure. Grey blocks represents data, blue blocks stands for analysis methods/models.}
    %     \label{fig:ChProp}
    % \end{figure}

@@

    The stage/discharge relationship is sensitive to sudden changes caused by morphogenic floods or other causes affecting the flow characteristics. Relying on residuals between the gaugings and rating curve is the most common approach to monitor the stability of this relationship over time. The method proposed by \citet{darienzo_detection_2021} is used in this work. First, a baseline rating curve is estimated from the whole gaugings dataset, using BaRatin method \citep{le_coz_combining_2014}. The residuals between gaugings and the rating curve are determined, and a statistical segmentation test is applied to them. This segmentation procedure accounts for the residuals uncertainty, coming from both the gaugings uncertainty and the rating curve uncertainty. The optimal number of gaugings is determined based on the Bayesian Information criterion (BIC). Then, the same steps are applied recursively to each sub-period determined earlier. The recursive procedure is stopped when the BIC indicates that a single period is optimal. The results are not only the dates of the rating shifts but the posterior probability density functions (pdf) of change point times. We assume affecting the shift time to the time of the maximum stage included in the posterior credibility interval. Prior knowledge is provided on the expected magnitude of the residuals, the number of segments at each iteration and the minimum time lag between two consecutive change points.

@@

    \paragraph{}
    Once the stability periods have been identified, the next step is to estimate the rating curve associated with each period. \citet{mansanarez_shift_2019} developed a stage-period-discharge (SPD) model "based on the physical interpretation of changes in the stage-discharge relation across a series of stability periods". The SPD model is based on the BaRatin model \citep{le_coz_combining_2014}.
    \paragraph{} BaRatin uses the Bayesian paradigm to estimate the parameters of the general rating curve equation, a combination of power equations: $Q = a(h-b)^c$, where $Q$ is the discharge in m\textsuperscript{3}/s, the flow depth is the difference between the stage $h$ (in m) and an offset $b$, and $a$ and $c$ being the coefficient and the exponent of the power function. The rating curve equation is deduced from a hydraulic analysis of the gauging station, aimed at identifying the main hydraulic controls governing the stage-discharge relation. The multiple controls can be activated successively or simultaneously. Bayesian inference allows deriving the posterior distribution of rating curve parameters by combining hydraulic information (priors for parameters of each hydraulic controls) and information from gaugings with uncertainty (likelihood).
    \paragraph{}
    The SPD model is estimating the rating curves of each stability based on the same principle, and is considering that several identified parameters are varying in time, throughout the stability periods. An important step is the identification of those varying parameters based on hydraulic analysis of the site. Generally, offsets or channel widths are suspected to change. The so-called local changes are supposed to affect the lowest control only (for instance the movement of the controlling riffle), whereas the global changes affect various controls at the same time (for instance, the scouring or filling of the channel, affecting low flows and main channel offsets at the same time). One way to follow the evolution of riverbed elevation and to help specifying priors for rating curve vertical evolution is to study the long-term evolution of the yearly lowest stages, as described by \citet{lapuszek_methods_2015}. See \citet{mansanarez_shift_2019} for a detailed description of prior specification for time-varying rating curves.

@@

    Many sources of error having distinct statistical properties can affect stage measurements, as described in \citet{horner_impact_2018}. Five different sources of error ($\delta_{1,...,5}$) affecting stage measurements are considered. Let $h(t)$ be the measured maximum stage of a day $t$. $\hbar(t)$, the unknown true maximum stage is assumed to be approximated by the following equation:

@@

    \paragraph{}Staff gauge reading errors $\delta_1 \sim \mathcal{N}(0,\sigma_1)$ originate from operators reading the gauge, where $\sigma_1$ depends on the resolution of the graduations (usually 1 cm), and can be increased by waves, especially during floods \citep{mcmillan_benchmarking_2012}.

    \paragraph{}Most stage measurements nowadays are done with automatic sensors of various types such as pressure sensors, floats, radars, and they require a calibration to link the water stage to the measured proxy (respectively the pressure of the water column, the height of a float, or the air draught). Two types of errors arise from this process: sensor errors $\delta_2 \sim \mathcal{N}(0,\sigma_2)$, where $\sigma_2$ being usually estimated by the sensor manufacturer, and sensor calibration errors $\delta_3 \sim \mathcal{N}(0,\sigma_3)$ that are related to the corrections made by operators when comparing the stage measured by the sensor to the actual stage at the staff gauge reference. An operator error at this step could affect the stage measurement until the next calibration. Sensor drifts are ignored, for instance, sensor calibration error $\delta_3$ is assumed constant between two calibrations and can be represented by drawing a new random value at each operator intervention.

    \paragraph{}Datum errors $\delta_4  \sim \mathcal{N}(0,\sigma_4)$ are related to changes in the datum reference elevation of the staff gauge zero value and possible discontinuity between successive gauges. Similarly to $\delta_3$, this error is constant between two gauge changes or datum reference measurements.

    \paragraph{}Measurement frequency errors $\delta_5$ are related to the inadequacy of the frequency of measurement with respect to the speed of stage variations. Unlike other types of stage errors, this error is hence necessarily positive, which calls for using a positive distribution such as the Log-normal or Exponential distribution at each measurement time step. The parameters of this distribution can be estimated with data from the recent period, by analyzing the difference between the daily maximum stage derived from the high-frequency sensor measurement and that from an infrequent fixed-time reading. Note that the measurement frequency error for hourly (or less) measurements are considered negligible when considering large rivers with slow variations.

    \paragraph{}
    To sum up, $\delta_1$, $\delta_2$ and $\delta_5$ errors are drawn at each time step, while $\delta_3$ and $\delta_4$ errors are only drawn at specific calibration times. $\delta_1$ to $\delta_4$ are assumed Gaussian with known standard deviations, while $\delta_5$ is assumed Exponential with parameter estimated by subsampling recent measurements. Monte Carlo methods are applied to combine these multiple stage uncertainty sources. The uncertainty around stage measurements from eq. \ref{eq:StageError} is represented by 500 possible realizations of the stage $h(t)$.

@@

   Stage measurements realizations can be propagated through uncertain rating curves, following the approach described by \citet{horner_impact_2018}. To estimate the contribution of the different sources of streamflow uncertainties, four scenarii are considered:

   \begin{itemize}
       \item \textbf{Case 1: Maxpost streamflow.} Stage measurements are free of uncertainties and are assumed to be the median of the stage time series realisations. Rating curve estimation is also free of uncertainties. The unique stage time series is propagated trough the maxpost (Maximum A Posteriori, i.e. the most probable) rating curves, resulting in one set of discharge time series. 
       \item \textbf{Case 2: Stage uncertainty.} Now, stage measurements are uncertain. This uncertainty is represented by $n$ possible stage time series, propagated through the maxpost rating curves. Thus, $n$ sets of discharge time series are obtained.
       \item \textbf{Case 3: Stage and parametric rating curve uncertainty.} Stage measurements are uncertain and rating curve estimation is affected by parametric uncertainties (represented by $m$ sets of rating curves) but is not affected by remnant uncertainties. The $n$ stage time series are propagated through $m$ sets of rating curves, leading to $n \times m$ sets of discharge time series. 
       \item \textbf{Case 4: Total streamflow uncertainty.} Stage measurements are uncertain and rating curve estimation is affected by both parametric and remnant uncertainties, represented by $m$ sets of rating curves. The $n$ stage time series are propagated through $m$ sets of rating curves, leading to $n \times m$ sets of discharge time series to which remnant errors are applied. This represents the total streamflow uncertainty.
   \end{itemize}

@@

    \subsubsection{Pont de Beaucaire (1816 - 1967)}
